Modeling trend in temperature volatility using generalized LASSO

### 2018-05-18 nips-2018/paper.tex
ready to submit

\author{Arash Khodadadi\\
 Department of Statistics\\
 Indiana University\\
 Bloomington, IN 47408 \\
 \texttt{arakhoda@indiana.edu}
\And
  Daniel J.\ McDonald \\
 Department of Statistics\\
 Indiana University\\
 Bloomington, IN 47408 \\
 \texttt{dajmcdon@indiana.cmu.edu}}

@@

\begin{abstract}
In this paper, we present methodology for estimating trends in
spatio-temporal volatility. We give two algorithms for computing our
estimator which are tailored for dense, gridded observations over both
space and time, though these can be easily extended to other
structures (time-varying network flows, neuroimaging). We motivate our
methodology using by applying it to a massive climate dataset and
discuss the implications for climate analysis.
\end{abstract}

@@

Trends in terrestrial temperature variability are perhaps more relavant for species
viability than trends in mean
temperature~\citep{huntingford_no_2013},
because, an increase in the
temperature variability will increase the probability of extreme hot
outliers~\citep{VasseurDeLong2014}. Recent climate literature
suggests that it is more difficult for society to adapt to these
extremes than to the gradual increase in the mean temperature
\citep{hansen_perception_2012,huntingford_no_2013}. Furthermore, the willingness of popular media to
emphasize the prevalence extreme cold events coupled with a
fundamental misunderstanding of the relationship between climate (the global
distribution of weather over the long run) and weather (observed
short-term, localized behavior) leads to public misunderstanding of climate
change. In fact, increased frequency of extreme cold events in the
northern hemisphere is can be partially attributed to increases in
mean climate but is also due to increases in temperature
variance~\citep{Screen2014,FischerBeyerle2013,TrenberthZhang2014}.

@@

These and other related research, e.g.,
\citep{rhines_frequent_2013}) have several shortcomings. First, no
statistical analysis has been performed to examine if the changes in
the SD are statistically significant. Second, the methodologies for
computing the SDs are highly sensitive to the choice of base period. Third,
and most importantly, temporal and spatial correlations between the
observations are ignored.

@@

As with~\eqref{eq:l1tf}, one can solve~\eqref{eq:l1tf_var} using the
PDIP algorithm (as in, e.g.,
\texttt{cvxopt}~\citep{andersen_cvxopt:_2013}). In each iteration of
PDIP we need to compute a search direction by taking a Newton
step on a system of nonlinear equations. Due to space limitations, we
defer details to Appendix B in the Supplement.

@@

At a specific time $t$, the data is measured on a grid of points with
$n_r$ rows and $n_c$ columns for a total of $S=n_r\times n_c$ spatial
locations. Let $y_{ijt}$ denote the value of the 
observation at time $t$ on the $i^\text{th}$ row and $j^\text{th}$
column of the grid, and $h_{ijt}$ denote the corresponding hidden
variable. We seek to impose both temporal and spatial smoothness
constraints on the hidden variables. Specifically, we seek a solution
for $h$ which is piecewise linear in time and piecewise constant in
space (although higher-order smoothness can be imposed with minimal
alterations to the methodology). We achieve this goal
by solving the following optimization problem: 
\begin{align}
\min_h &\sum_{i,j,t}h_{ijt}+y_{ijt}^2e^{-h_{ijt}}
+\lambda_1 \sum_{i,j} \sum_{t=1}^{T-2} \left|h_{ijt}-2h_{ij(t+1)}+h_{ij(t+2)}\right|\\
&\quad+\lambda_2 \sum_{t,j} \sum_{i=1}^{n_r-1} \left|h_{ijt}-h_{(i+1)jt}\right|
+\lambda_2 \sum_{t,i} \sum_{j=1}^{n_c-1} \left|h_{ijt}-h_{i(j+1)t}\right|
\label{eq:l1tf_var_st}
\end{align}

The first term in the objective is proportional to the negative
log-likelihood, the second is the temporal penalty for the
time-series at each location $(i,j)$, while the third and fourth,
penalize the difference between the estimated variance of two
vertically and horizontally adjacent points, respectively. The spatial
component of this
penalty is a special case of trend
filtering on graphs~\citep{WangSharpnack2016} which penalizes the difference
between the estimated values of the signal on the connected
nodes. As before, we can write
\eqref{eq:l1tf_var_st} in matrix form where $h$ is an $T\times
S$ vector and $D$ is replaced by $D_{ST} \in \mathbb{R}^{(N_t+N_s)
  \times (T \cdot S)	}$, where $N_t=S \cdot (T-2)$ and $N_s=T \cdot
(2n_rn_c-n_r)$ are the number of temporal and spatial constraints,
respectively\footnote{$N_s$ is obtained by counting the number of
  unique constraints at each location and at all times.}. Then, as we 
have two different tuning parameters for the temporal and spatial
components, we write $\Lambda =\left[\lambda_1\one_{N_t}^\top,\;
  \lambda_2\one_{N_s}^\top\right]^\top$ leading
to:\footnote{Throughout the paper, we use $|x|$ for both scalars and
  vectors. For vectors we use this to denote a vector obtained by
  taking the absolute value of each entry of $x$.} 
% let $h$ be a
% vector whose first $T$ entries are $h_{11t}$ for $t=1,...,T$, its next
% $T$ entries are $h_{21t}$ and so on, then the optimization problem in
% the matrix form is as
\begin{equation}
\min_h -l(y\given h)+ \Lambda^\top | D_{ST}h |.
\label{eq:l1tf_var_st_mat}
\end{equation}
\attn{equations should have proper punctuation as if they were a sentence.}
% where $\Lambda = \mathbf{I}_{Tn_rn_c}\left[\lambda_1\one_T^\top,\;
%   \lambda_2\one_{n_rn_c}^\top\right]$ and 
% $D_{total}^t=[D_{t}^t|D_{s}^t]$, where $n_t$ and $n_s$ are the number
% of rows of $D_{t}$ and $D_{s}$, respectively (see below). The matrix
% $D_t$ is the following block-diagonal matrix and corresponds to the
% temporal penalty: 
% \begin{equation}
% D_t=\begin{bmatrix}
% D &  & \\ 
%  & \ddots & \\
%  &  & D
% \end{bmatrix}
% \label{eq:d_t_matrix}
% \end{equation}
%  where $D$ was defined in \eqref{eq:d_matrix}. The number of the
%  diagonal blocks is equal to the grid size $n_r \times n_c$. Each row
%  of the matrix $D_s$ corresponds to one spatial constraint in
%  \eqref{eq:l1tf_var_st}. For example, the first $T$ rows correspond to
%  $|h_{11t}-h_{21t}|$ for $t=1,...,T$, the next $T$ rows correspond to
%  $|h_{11t}-h_{12t}|$, and so on. 
% \attn{Detailed discussion of the PDIP to here}

% \attn{Add back the full form of $D$ above? It was too cumbersome
%   before, but possibly necessary}

@@

For a spatial grid of size $S$ and $T$ time steps, $D_{ST}$ will have
$3Tn_rn_c-2n_rn_c-Tn_r$ rows and $ST$ columns. For a $1^\circ\times
1^\circ$ grid over the entire northern hemisphere and daily data over
10 years, we have $n_r=90$, $n_c=180$, $T=3650$ and so $D_{ST}$ has
approximately $10^8$ columns and $10^8$ rows. In each step of the PDIP
algorithm, we need to solve a linear system of equations in $A$ which
depends on $D_{ST}^\top D_{ST}$ (see appendix A and B). Therefore,
applying the PDIP directly is infeasible for our data.\footnote{We
  note that this is a highly structured and sparse matrix, but, unlike
  trend filtering alone, it is not banded. We are unaware of general
  linear algebra techniques for inverting such matrix, despite our
  best efforts.}

@@

Given an optimization problem of the form $\min_h f(h)$, where
$h\in\mathbb{R}^n$ is the global variable and
$f(h):\mathbb{R}^n \rightarrow \mathbb{R}\cup \{+\infty\}$ is
convex. Consensus optimization breaks this problem
into several smaller sub-problems that can be solved independently in
each iteration of optimization.

Assume it is possible to define a set of local variables
$x_i \in \mathbb{R}^{n_i}$ such that $f(h)=\sum_i f_i(x_i)$, where
each $x_i$ is a subset of the global variable $h$. More specifically,
each entry of the local variables corresponds to an entry of the
global variable. Therefore we can define a mapping $\mathscr{G}(i,j)$
from the local variable indices into the global variable indices:
$k=\mathscr{G}(i,j)$ means that the $j^\text{th}$ entry of $x_i$ is
$h_k$ (or $(x_i)_j=h_k$). For ease of notation, define $\tilde{h}_i
\in \mathbb{R}^{n_i}$ as $(\tilde{h}_i)_j=h_{\mathscr{G}(i,j)}$. Then,
the original optimization problem is equivalent to the following
problem:  
\begin{equation}
\begin{aligned}
\min_{\{x_1,...,x_N  \}} \sum_i f_i(x_i)\\
 s.t. \,\,\, \tilde{h}_i=x_i. \,\,\,\,\,\,
\end{aligned}
\label{eq:consADMM}
\end{equation}
It is important to note that each entry of the global variable may
correspond to several entries of the local variables and so the
constraints $\tilde{h}_i=x_i$ enforce the consensus between the local
variables corresponding to the same global variable.  
The augmented Lagrangian corresponding to 
\eqref{eq:consADMM} is $L_\rho(x,h,y)=\sum_i
\big(f_i(x_i)+u_i^\top(x_i-\tilde{h}_i) + (\rho/2) \lVert
x_i-\tilde{h}_i \lVert_2^2 \big)$. Now, we can apply ADMM to $L_\rho$.
% which results in the following ADMM updates:
% \begin{align}
% x_i&\leftarrow\argmin_{x_i} f_i(x_i) + (u_i)^\top x_i +
%      (\rho/2)  \norm{ x_i-\tilde{h}_i }_2^2\\
% h_k&\leftarrow (1/S_k)\sum_{\mathscr{G}(i,j)=k} (x_i)_j \label{eq:ADMM_steps}\\
% u_i&\leftarrow u_i + \rho (x_i-\tilde{h}_i).
% \end{align}
% \attn{If you hate the left arrow, that's fine. But change it
%   everywhere, and use $=$. The $:=$ notation means that ``thing on
%   left is defined to be thing on right'' and isn't really appropriate here.}
% Here, $S_k$ is the number of local variable entries that correspond to
% $h_k$, and $u_i$ are the Lagrange multipliers.  
To solve the optimization problem \eqref{eq:l1tf_var_st_mat} using
this method, we need to address two 
questions: first, how to choose the 
local variables $x_i$, and second, how to the update them.
\begin{figure}[tb]
  \centering
  \includegraphics[height=.2\textheight]{Figures/data_cube}
  \caption{The cube represents the global variable $h$ in space and
    time. The sub-cubes specified by the white lines are
    $x_i$.}
  \label{fig:data_cube}
\end{figure}

In \autoref{fig:data_cube}, the global variable $h$ is represented
as a cube (using the subset of the US as an example). We decompose $h$
into sub-cubes as shown by white lines. With
this definition of $x_i$, the objective
\eqref{eq:l1tf_var_st_mat} decomposes as $\sum_i f_i(x_i)$ where
$f_i(x_i)=-l(y_i\given x_i)+\Lambda_{(i)}^\top |D_{(i)}x_i|$, and
$\Lambda_{(i)}$ and $D_{(i)}$ contain the temporal and spatial
penalties corresponding to $x_i$ only. 
% The $x$-update step in
% \eqref{eq:ADMM_steps} is the following optimization problem:
% $x_i^{m+1}:=\argmin_{x_i} ( f_i(x_i) + 
% (u_i^m)^\top x_i + (\rho/2) \lVert x_i-\tilde{z}_i^m \lVert_2^2)$.
Thus, with this choice of the local variables $x_i$, we solve the
$x$-update using the PDIP method. 
% As it 
% was explained in \autoref{sec:l1tf_var}, to use PDIP we need to
% compute the dual problem, which in turn needs the computation of the
% Fenchel conjugate of the loss function. In addition, referring to
% \autoref{alg:pdip}, we need to compute the gradient and Jacobian of
% the conjugate functions. It is important to note that compare to the
% optimization problem \autoref{eq:l1tf_var}, the loss function in the
% x-update step in \autoref{eq:ADMM_steps} includes the quadratic term
% $(\rho/2) \lVert x_i-\tilde{z}_i^m \lVert_2^2$. This makes the
% computation more involved than \autoref{sec:l1tf_var}. 
\autoref{alg:conADMM} gives the general version of this algorithm. A
more detailed discussion of the is in the Supplement.

@@

Consider the generic optimization problem
$\min_x f(x)+g(Dx)$
where $x\in \mathbb{R}^n$ and $D\in \mathbb{R}^{m\times n}$. Each
iteration of the linearized ADMM
algorithm~\citep{parikh_proximal_2014} for solving this problem 
has the form
\begin{align}
x & \leftarrow \prox_{\mu f} \left(x - (\mu/\rho)D^\top (D x - z + u )\right)\\
z & \leftarrow \prox_{\rho g} \left(z + u\right) \label{eq:linADMM_steps}\\
u & \leftarrow u + D x - z
\end{align}
where the algorithm parameters $\mu$ and $\rho$ satisfy $0 < \mu < \rho/\norm{D}_2^2$, $z,u\in \mathbb{R}^m$ and the proximal operator is defined as
$
 \prox_{\alpha f}(u) = \min_x \,\, \alpha \cdot f(x)+\frac{1}{2} \norm{ x-u}_2^2.
$

%Clearly, \eqref{eq:l1tf_var_st} has this form necessary for using this algorithm.
% The optimization problem 
% form \eqref{eq:linADMM_opt} by defining
% \begin{align}
% & f(x):= \sum_{k} f_k(x_k) := \sum_{k}x_k+y_{k}^2e^{-x_{k}}\\
% & g(z):= \sum_{l} g_l(z_l) := \sum_{l} \lambda_l |z_l|\label{eq:linADMM_fg}\\
% & z = Dx
% \end{align}
% where $y_k$ is the $k^{th}$ entry of the vector whose entries are
% $y_{ijt}$, and $\lambda_l$ is the $l^{th}$ entry of the vector
% $\Lambda^\top=(\lambda_t\textbf{e}_{n_t}^\top|\lambda_s\textbf{e}_{n_s}^\top)$
% (see \autoref{sec:exten}). 
% To perform the steps in \eqref{eq:linADMM_steps}, we need to evaluate
% $\prox_{\mu f}$ and $\prox_{\rho g}$. 
Proximal
algorithms are feasible when these proximal operators can be
evaluated efficiently which, as we show next, is the case for our
problem.

@@

We estimated the linearized ADMM for all combinations of values of
$\lambda_t$ and $\lambda_s$ from the sets $\lambda_t \in
\{0,1,5,10,50,100\}$ and $\lambda_s \in \{0,0.05,0.1,0.2,0.3\}$. For
each pair, we then compute the mean absolute error (MAE) between the
estimated variance and the true variance at all locations and all time
steps. For $\lambda_t=5$ and $\lambda_s=0.1$, the MAE was minimized. The
left panel of \autoref{fig:true_fitted_var} shows the true and the
estimated standard deviation at location (0,0) using $\lambda_s=0.1$
and $\lambda_t=5$ (blue) and $\lambda_t=100$ (green). As we can see,
larger than optimal value of $\lambda_t$ leads to estimated values
which are ``too smooth".

@@

The middle panel of \autoref{fig:true_fitted_var} shows the
convergence of both methods. Each iteration of the linearized
algorithm takes 0.01 seconds on average while each iteration of the
consensus ADMM takes about 20 seconds.

To further examine the performance of the proposed model, we next
compare it to three alternatives: a model which does not consider the
spatial smoothness (equivalent to fitting the model in
\autoref{sec:l1tf_var} to each time-series separately), a model which
does not consider imposes only spatial smoothness, and a GARCH(1,1) model. We
simulated 100 datasets using the method explained above with $\sigma_s
\sim \mathsf{uniform}(4,7)$. The right panel of
\autoref{fig:true_fitted_var} shows the boxplot of the MAE for these
models. Interestingly, the proposed model with optimal parameters
outperforms GARCH(1,1) in estimating the true value of the variance.

@@

Consensus ADMM in~\autoref{sec:consOpt} is appropriate only if
we parallelize it over multiple machines, and it is significantly
slower on our simulated data, so we do not pursue it
further here. All the results reported in this section are obtained
using \autoref{alg:linADMM}. We
applied this algorithm to the northern hemisphere of the ERA-20C
dataset available from the \href {https://www.ecmwf.int}{European Center for Medium-Range
  Weather Forecasts}. The data are
the 2 meter temperature measured daily at 12 p.m from January 1, 1960
to December 24, 2010.

The Supplement explains some preprocessing and investigates some
properties of the time-series of different locations on
earth. \autoref{fig:bloom_estimatedSD} a processed time-series for a
single location.
The variance of this time-series has an irregular cyclic
behavior. Additionally, the time-series of other locations show 
different patterns. These observations motivated the need to develop a
non-parametric framework for this
problem. \autoref{fig:bloom_estimatedSD} also shows the
estimated SD obtained using the method of
\autoref{sec:l1tf_var}. 
% Since the estimated SD captures the periodic
% behavior of 
% volatility, it is hard to distinguish any long-term trend. Therefore,
% we compute the annual average of these 
% estimates. However, as this figure shows, the annual trend is not
% smooth. This is because in the optimization problem
% \eqref{eq:l1tf_var}, the smoothness of the annual trend is not
% encouraged. Therefore, in Appendix D we propose a simple method to
% encourage the smoothness of the annual average trend. The estimated
% SDs using this method is shown in the right panel of
% \autoref{fig:bloom_estimatedSD}. The annual average of the estimated
% SDs shows a linear trend with a positive slope.

@@

\paragraph{Model selection}
One common method for choosing the penalty parameters in the Lasso
problems is to find the solution for a range of the values of these
parameters and then choose the values which minimize a model selection
criterion. However, such analyses needs the computation of the degrees
of freedom. Several previous work have investigated the df in
generalized lasso problems
\citep{tibshirani_degrees_2012,hu_dual_2015,zeng_geometry_2017}. However,
all these studies have considered the linear regression problem and,
to the best of our knowledge, the problem of computing the df for
generalized lasso with general objective function has not been
considered yet.

In this paper, we use a heuristic method for choosing $\lambda_t$ and
$\lambda_s$: we compute the optimal solution for a range of values of
these parameters and choose the values which minimize
$\mathscr{L}(\lambda_t,\lambda_s)=-l(y|h)+ \sum \lVert D_{total}h
\lVert$. This objective is a compromise between the negative log
likelihood ($-l(y|h)$) and the complexity of the solution ($\sum
\lVert D_{total}h \lVert$). For smoother solutions the value of $\sum
\lVert D_{total}h \lVert$ will be smaller but with the cost of larger
$-l(y|h)$.

We computed the optimal solution for all the combinations of the
following sets of values: $\lambda_t \in \{0,2,4,8,10,15,200,1000\} \, \, ,
\lambda_s \in \{0,.1,.5,2,5,10\}$. The best combination was
$\lambda_t=4$ and $\lambda_s=2$. All the analyses in the next section
are performed on the solution for these values.

@@

\bibliographystyle{abbrvnat}
\small
\bibliography{nips-references}

>>>>

\author{Arash Khodadadi\\
 Department of Statistics\\
 Indiana University\\
 Bloomington, IN 47408 \\
 \texttt{arakhoda@indiana.edu}
\And
  Daniel J.\ McDonald \\
 Department of Statistics\\
 Indiana University\\
 Bloomington, IN 47408 \\
 \texttt{dajmcdon@indiana.edu}}

@@

\begin{abstract}
In this paper, we present methodology for estimating trends in
spatio-temporal volatility. We give two algorithms for computing our
estimator which are tailored for dense, gridded observations over both
space and time, though these can be easily extended to other
structures (time-varying network flows, neuroimaging). We motivate our
methodology by applying it to a massive climate dataset and
discuss the implications for climate analysis.
\end{abstract}

@@

Trends in terrestrial temperature variability are perhaps more relavant for species
viability than trends in mean
temperature~\citep{huntingford_no_2013},
because, an increase in the
temperature variability will increase the probability of extreme hot
outliers~\citep{VasseurDeLong2014}. Recent climate literature
suggests that it is more difficult for society to adapt to these
extremes than to the gradual increase in the mean temperature
\citep{hansen_perception_2012,huntingford_no_2013}. Furthermore, the willingness of popular media to
emphasize the prevalence extreme cold events coupled with a
fundamental misunderstanding of the relationship between climate (the global
distribution of weather over the long run) and weather (observed
short-term, localized behavior) leads to public misunderstanding of climate
change. In fact, increased frequency of extreme cold events in the
northern hemisphere can be partially attributed to increases in
mean climate but is also due to increases in temperature
variance~\citep{Screen2014,FischerBeyerle2013,TrenberthZhang2014}.

@@

These and other related research, e.g.,
\citep{rhines_frequent_2013}, have several shortcomings. First, no
statistical analysis has been performed to examine if the changes in
the SD are statistically significant. Second, the methodologies for
computing the SDs are highly sensitive to the choice of base period. Third,
and most importantly, temporal and spatial correlations between the
observations are ignored.

@@

As with~\eqref{eq:l1tf}, one can solve~\eqref{eq:l1tf_var} using the
PDIP algorithm (as in, e.g.,
\texttt{cvxopt}~\citep{andersen_cvxopt:_2013}). In each iteration of
PDIP we need to compute a search direction by taking a Newton
step on a system of nonlinear equations. Due to space limitations, we
defer details to Appendix A in the Supplement, where we show how to derive the dual of this optimization problem and compute the first and second derivatives of the dual objective function.

@@

At a specific time $t$, the data is measured on a grid of points with
$n_r$ rows and $n_c$ columns for a total of $S=n_r\times n_c$ spatial
locations. Let $y_{ijt}$ denote the value of the 
observation at time $t$ on the $i^\text{th}$ row and $j^\text{th}$
column of the grid, and $h_{ijt}$ denote the corresponding hidden
variable. We seek to impose both temporal and spatial smoothness
constraints on the hidden variables. Specifically, we seek a solution
for $h$ which is piecewise linear in time and piecewise constant in
space (although higher-order smoothness can be imposed with minimal
alterations to the methodology). We achieve this goal
by solving the following optimization problem: 
\begin{align}
\min_h &\sum_{i,j,t}h_{ijt}+y_{ijt}^2e^{-h_{ijt}}
+\lambda_1 \sum_{i,j} \sum_{t=1}^{T-2} \left|h_{ijt}-2h_{ij(t+1)}+h_{ij(t+2)}\right|\\
&\quad+\lambda_2 \sum_{t,j} \sum_{i=1}^{n_r-1} \left|h_{ijt}-h_{(i+1)jt}\right|
+\lambda_2 \sum_{t,i} \sum_{j=1}^{n_c-1} \left|h_{ijt}-h_{i(j+1)t}\right|
\label{eq:l1tf_var_st}.
\end{align}

The first term in the objective is proportional to the negative
log-likelihood, the second is the temporal penalty for the
time-series at each location $(i,j)$, while the third and fourth,
penalize the difference between the estimated variance of two
vertically and horizontally adjacent points, respectively. The spatial
component of this
penalty is a special case of trend filtering on graphs~\citep{WangSharpnack2016} which penalizes the difference between the estimated values of the signal on the connected nodes. As before, we can write \eqref{eq:l1tf_var_st} in matrix form where $h$ is an $T\times S$ vector and $D$ is replaced by $D_{ST} \in \mathbb{R}^{(N_t+N_s) \times (T \cdot S)	}$, where $N_t=S \cdot (T-2)$ and $N_s=T \cdot (2n_rn_c-n_r)$ are the number of temporal and spatial constraints, respectively\footnote{$N_s$ is obtained by counting the number of unique constraints at each location and at all times.}. The exact form of this matrix in clarified in Appendix A in the Supplement. Then, as we have two different tuning parameters for the temporal and spatial components, we write $\Lambda =\left[\lambda_1\one_{N_t}^\top,\;  \lambda_2\one_{N_s}^\top\right]^\top$ leading to:\footnote{Throughout the paper, we use $|x|$ for both scalars and vectors. For vectors we use this to denote a vector obtained by taking the absolute value of each entry of $x$.} 
% let $h$ be a
% vector whose first $T$ entries are $h_{11t}$ for $t=1,...,T$, its next
% $T$ entries are $h_{21t}$ and so on, then the optimization problem in
% the matrix form is as
\begin{equation}
\min_h -l(y\given h)+ \Lambda^\top | D_{ST}h |.
\label{eq:l1tf_var_st_mat}
\end{equation}

@@

For a spatial grid of size $S$ and $T$ time steps, $D_{ST}$ will have
$3Tn_rn_c-2n_rn_c-Tn_r$ rows and $ST$ columns. For a $1^\circ\times
1^\circ$ grid over the entire northern hemisphere and daily data over
10 years, we have $n_r=90$, $n_c=360$, $T=3650$ and so $D_{ST}$ has
approximately $10^8$ columns and $10^8$ rows. In each step of the PDIP
algorithm, we need to solve a linear system of equations which
depends on $D_{ST}^\top D_{ST}$ (see appendix A and B). Therefore,
applying the PDIP directly is infeasible for our data.\footnote{We
  note that this is a highly structured and sparse matrix, but, unlike
  trend filtering alone, it is not banded. We are unaware of general
  linear algebra techniques for inverting such matrix, despite our
  best efforts.}

@@

Consider an optimization problem of the form $\min_h f(h)$, where
$h\in\mathbb{R}^n$ is the \textit{global variable} and
$f(h):\mathbb{R}^n \rightarrow \mathbb{R}\cup \{+\infty\}$ is
convex. Consensus optimization breaks this problem
into several smaller sub-problems that can be solved independently in
each iteration of optimization.

Assume it is possible to define a set of local variables
$x_i \in \mathbb{R}^{n_i}$ such that $f(h)=\sum_i f_i(x_i)$, where
each $x_i$ is a subset of the global variable $h$. More specifically,
each entry of the local variables corresponds to an entry of the
global variable. Therefore we can define a mapping $\mathscr{G}(i,j)$
from the local variable indices into the global variable indices:
$k=\mathscr{G}(i,j)$ means that the $j^\text{th}$ entry of $x_i$ is
$h_k$ (or $(x_i)_j=h_k$). For ease of notation, define $\tilde{h}_i
\in \mathbb{R}^{n_i}$ as $(\tilde{h}_i)_j=h_{\mathscr{G}(i,j)}$. Then,
the original optimization problem is equivalent to the following
problem:  
\begin{equation}
\begin{aligned}
\min_{\{x_1,...,x_N  \}} \sum_i f_i(x_i)\\
 s.t. \,\,\, \tilde{h}_i=x_i. \,\,\,\,\,\,
\end{aligned}
\label{eq:consADMM}
\end{equation}
It is important to note that each entry of the global variable may
correspond to several entries of the local variables and so the
constraints $\tilde{h}_i=x_i$ enforce the consensus between the local
variables corresponding to the same global variable.  
The \textit{augmented Lagrangian} corresponding to 
\eqref{eq:consADMM} is $L_\rho(x,h,y)=\sum_i
\big(f_i(x_i)+u_i^\top(x_i-\tilde{h}_i) + (\rho/2) \lVert
x_i-\tilde{h}_i \lVert_2^2 \big)$. Now, we can apply ADMM to $L_\rho$. This results in solving $N$ independent optimization problems followed by a step to achieve consensus among the solutions in each iteration.
% which results in the following ADMM updates:
% \begin{align}
% x_i&\leftarrow\argmin_{x_i} f_i(x_i) + (u_i)^\top x_i +
%      (\rho/2)  \norm{ x_i-\tilde{h}_i }_2^2\\
% h_k&\leftarrow (1/S_k)\sum_{\mathscr{G}(i,j)=k} (x_i)_j \label{eq:ADMM_steps}\\
% u_i&\leftarrow u_i + \rho (x_i-\tilde{h}_i).
% \end{align}
% \attn{If you hate the left arrow, that's fine. But change it
%   everywhere, and use $=$. The $:=$ notation means that ``thing on
%   left is defined to be thing on right'' and isn't really appropriate here.}
% Here, $S_k$ is the number of local variable entries that correspond to
% $h_k$, and $u_i$ are the Lagrange multipliers.  
To solve the optimization problem \eqref{eq:l1tf_var_st_mat} using
this method, we need to address two 
questions: first, how to choose the 
local variables $x_i$, and second, how to the update them.
\begin{figure}[tb]
  \centering
  \includegraphics[height=.13\textheight]{Figures/data_cube}
  \caption{The cube represents the global variable $h$ in space and
    time. The sub-cubes specified by the white lines are
    $x_i$.}
  \label{fig:data_cube}
\end{figure}

In \autoref{fig:data_cube}, the global variable $h$ is represented
as a cube. We decompose $h$
into sub-cubes as shown by white lines. Each local variable inside the sub-cubes corresponds to only one global variable. The local variables on the border (white lines), however, correspond to more than one global variable. With
this definition of $x_i$, the objective
\eqref{eq:l1tf_var_st_mat} decomposes as $\sum_i f_i(x_i)$ where
$f_i(x_i)=-l(y_i\given x_i)+\Lambda_{(i)}^\top |D_{(i)}x_i|$, and
$\Lambda_{(i)}$ and $D_{(i)}$ contain the temporal and spatial
penalties corresponding to $x_i$ only in one sub-cube. Thus, we now need to use PDIP to solve $N$ problems each of size $n_i$, which is feasible for small enough $n_i$. \autoref{alg:conADMM} gives the general version of this algorithm. A more detailed discussion of this is in Appendix B of the Supplement where we show how to compute the dual and the derivatives of the augmented Lagrangian.

@@

Consider the generic optimization problem
$\min_x f(x)+g(Dx)$
where $x\in \mathbb{R}^n$ and $D\in \mathbb{R}^{m\times n}$. Each
iteration of the linearized ADMM
algorithm~\citep{parikh_proximal_2014} for solving this problem 
has the form
\begin{align}
x & \leftarrow \prox_{\mu f} \left(x - (\mu/\rho)D^\top (D x - z + u )\right)\\
z & \leftarrow \prox_{\rho g} \left(z + u\right) \label{eq:linADMM_steps}\\
u & \leftarrow u + D x - z
\end{align}
where the algorithm parameters $\mu$ and $\rho$ satisfy $0 < \mu < \rho/\norm{D}_2^2$, $z,u\in \mathbb{R}^m$ and the proximal operator is defined as $\prox_{\alpha f}(u) = \min_x \,\, \alpha \cdot f(x)+\frac{1}{2} \norm{ x-u}_2^2$. Proximal algorithms are feasible when these proximal operators can be evaluated efficiently which, as we show next, is the case for our problem.

%Clearly, \eqref{eq:l1tf_var_st} has this form necessary for using this algorithm.
% The optimization problem 
% form \eqref{eq:linADMM_opt} by defining
% \begin{align}
% & f(x):= \sum_{k} f_k(x_k) := \sum_{k}x_k+y_{k}^2e^{-x_{k}}\\
% & g(z):= \sum_{l} g_l(z_l) := \sum_{l} \lambda_l |z_l|\label{eq:linADMM_fg}\\
% & z = Dx
% \end{align}
% where $y_k$ is the $k^{th}$ entry of the vector whose entries are
% $y_{ijt}$, and $\lambda_l$ is the $l^{th}$ entry of the vector
% $\Lambda^\top=(\lambda_t\textbf{e}_{n_t}^\top|\lambda_s\textbf{e}_{n_s}^\top)$
% (see \autoref{sec:exten}). 
% To perform the steps in \eqref{eq:linADMM_steps}, we need to evaluate
% $\prox_{\mu f}$ and $\prox_{\rho g}$.

@@

We estimated the linearized ADMM for all combinations of values of
$\lambda_t$ and $\lambda_s$ from the sets $\lambda_t \in
\{0,1,5,10,50,100\}$ and $\lambda_s \in \{0,0.05,0.1,0.2,0.3\}$. For
each pair, we then compute the mean absolute error (MAE) between the
estimated variance and the true variance at all locations and all time
steps. For $\lambda_t=5$ and $\lambda_s=0.1$, the MAE was minimized. The
left panel of \autoref{fig:true_fitted_var} shows the true and the
estimated standard deviation at location (0,0) using $\lambda_s=0.1$
and $\lambda_t=5$ (blue) and $\lambda_t=100$ (green). As we can see,
larger than optimal value of $\lambda_t$ leads to estimated values
which are ``too smooth". The middle panel of \autoref{fig:true_fitted_var} shows the
convergence of both methods. Each iteration of the linearized
algorithm takes 0.01 seconds on average while each iteration of the
consensus ADMM takes about 20 seconds.

@@

To further examine the performance of the proposed model, we next
compare it to three alternatives: a model which does not consider the
spatial smoothness (equivalent to fitting the model in
\autoref{sec:l1tf_var} to each time-series separately), a model which only imposes spatial smoothness, and a GARCH(1,1) model. We
simulated 100 datasets using the method explained above with $\sigma_s
\sim \mathsf{uniform}(4,7)$. The right panel of
\autoref{fig:true_fitted_var} shows the boxplot of the MAE for these
models. Interestingly, the proposed model with optimal parameters
outperforms GARCH(1,1) in estimating the true value of the variance.

@@

Consensus ADMM in~\autoref{sec:consOpt} is appropriate only if
we parallelize it over multiple machines, and it is significantly
slower on our simulated data, so we do not pursue it
further here. All the results reported in this section are obtained
using \autoref{alg:linADMM}. We
applied this algorithm to the northern hemisphere of the ERA-20C
dataset available from the \href {https://www.ecmwf.int}{European Center for Medium-Range
  Weather Forecasts}. The data are
the 2 meter temperature measured daily at 12 p.m from January 1, 1960
to December 24, 2010. The Supplement explains some preprocessing and investigates some
properties of the time-series of different locations on
earth. \autoref{fig:bloom_estimatedSD} shows a processed time-series for a
single location.
The variance of this time-series has an irregular cyclic
behavior. Additionally, the time-series of other locations show 
different patterns. These observations motivated the need to develop a
non-parametric framework for this
problem. \autoref{fig:bloom_estimatedSD} also shows the
estimated SD obtained using the method of
\autoref{sec:l1tf_var}. 
% Since the estimated SD captures the periodic
% behavior of 
% volatility, it is hard to distinguish any long-term trend. Therefore,
% we compute the annual average of these 
% estimates. However, as this figure shows, the annual trend is not
% smooth. This is because in the optimization problem
% \eqref{eq:l1tf_var}, the smoothness of the annual trend is not
% encouraged. Therefore, in Appendix D we propose a simple method to
% encourage the smoothness of the annual average trend. The estimated
% SDs using this method is shown in the right panel of
% \autoref{fig:bloom_estimatedSD}. The annual average of the estimated
% SDs shows a linear trend with a positive slope.

@@

\paragraph{Model selection}
One common method for choosing the penalty parameters in the Lasso
problems is to find the solution for a range of the values of these
parameters and then choose the values which minimize a model selection
criterion. However, such analysis needs the computation of the degrees
of freedom. Several previous work have investigated the df in
generalized lasso problems
\citep{tibshirani_degrees_2012,hu_dual_2015,zeng_geometry_2017}. However,
all these studies have considered the linear regression problem and,
to the best of our knowledge, the problem of computing the df for
generalized lasso with general objective function has not been
considered yet. In this paper, we use a heuristic method for choosing $\lambda_t$ and
$\lambda_s$: we compute the optimal solution for a range of values of
these parameters and choose the values which minimize
$\mathscr{L}(\lambda_t,\lambda_s)=-l(y|h)+ \sum \lVert D_{total}h
\lVert$. This objective is a compromise between the negative log
likelihood ($-l(y|h)$) and the complexity of the solution ($\sum
\lVert D_{total}h \lVert$). For smoother solutions the value of $\sum
\lVert D_{total}h \lVert$ will be smaller but with the cost of larger
$-l(y|h)$. We computed the optimal solution for all the combinations of the
following sets of values: $\lambda_t \in \{0,2,4,8,10,15,200,1000\} \, \, ,
\lambda_s \in \{0,.1,.5,2,5,10\}$. The best combination was
$\lambda_t=4$ and $\lambda_s=2$. All the analyses in the next section
are performed on the solution for these values.

@@

\newpage
\bibliographystyle{abbrvnat}
\small
\bibliography{nips-references}
